On multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2022-07-29 paper/mcd2.tex
clean up nodal (NCP) residual presentation, and stopping criterion

During one application of \pr{nmcd-vcycle}, storage must be allocated for $\chi^j$, $\phi^j$, $g^j$, $y^j$, and $\ell^j$ on each level.  However, line 3 is essentially notational in that the finest-level iterate $g^J$ is the same as $w^J$; no copy is needed.  Similarly, in algorithm lines 13 and 15 no allocation is required because the $y^j$ space can be reused for $z^j$ as the coarser corrections are prolonged and accumulated.  Therefore the total storage of \pr{nmcd-vcycle}, excluding any in the smoothers/solver, but including for the original obstacle $\gamma^J$, is about $8 m^J$ real numbers\footnote{That is: $8\approx 5(4/3) + 1$.} under the standard estimate that a 2D, uniform-mesh V-cycle uses $4/3$ the memory of the finest level alone \cite[Section 2.4]{Trottenbergetal2001}.  For comparison, a single-level method needs at least $3 m^J$ storage for the vectors $\ell^J,\gamma^J,w^J$.

Regarding notation and semantics for the smoothers, note they may be iterated \id{down} and \id{up} times, respectively.  Both the smoothers and the coarsest-level solver are supposed to do in-place modifications of their final arguments, and of course they are in-exact solvers.  They must not modify their first three arguments.

Furthermore, repeated application of \pr{nmcd-vcycle}, or of the smoothers and solver at the levels, will not reduce the classical residual $f^J(u^J) - \ell^J$ to zero everywhere.  Recall that a finite-dimensional VI problem is equivalent to a \emph{nonlinear complementarity problem} (NCP) \cite{FacchineiPang2003}.  For example, corresponding to \eqref{eq:fe:vi} one could write
\begin{equation}
u^J \ge \gamma^J, \qquad f^J(u^J) \ge \ell^J, \qquad (u^J - \gamma^J)\left(f^J(u^J) - \ell^J\right) = 0.  \label{eq:fe:ncp}
\end{equation}
Thus, at convergence we only expect $f^J(u^J) = \ell^J$ at those locations where the constraint is inactive, where $u^J > \gamma^J$.  Therefore, recalling that a $j$th-level hat function based at node $x_p^j$ is denoted $\psi_p^j$, for an iterate $w^j$ define the \emph{downward nodal residual} by
\begin{equation}
(\hat r_{\text{down}}^j(w))_p = \begin{cases} \ip{f^j(w^j)-\ell^j}{\psi_p^j}, & w^j_p > \phi^j_p, \\
                                  \min\{\ip{f^j(w^j)-\ell^j}{\psi_p^j},0\}, & w^j_p = \phi^j_p, \end{cases} \label{eq:cpresidual}
\end{equation}
where $w^j_p,\phi^j_p$ denote the nodal values of $w^j,\phi^j$ as $x_p^j$, respectively.  The \emph{upward nodal residual} $\hat r_{\text{up}}(w)$ is similarly defined using $\chi^j$.

The nodal residuals $\hat r_{\text{down}}(w^j),\hat r_{\text{up}}(w^j)$ are nonzero if there are nodes where the NCP \eqref{eq:fe:ncp} is violated, either because $f^j(w^j)-\ell^j$ remains nonzero, in a weak sense, on the inactive set or because the same quantity is negative on the active set.  If $w^j$ is admissible, for instance if $w^j = g^j+z^j$ for $z^j \in \mathcal{U}^j$, inside \pr{nmcd-vcycle}, and if $\|\hat r_{\text{up}}(w^j)\|$ is small in some appropriate norm then we regard the $j$th-level problem as being numerically solved.  This provides stopping criteria for application of \pr{solver} on the coarsest mesh, and for terminating cycles of \pr{nmcd-vcycle} when addressing the original VI \eqref{eq:fe:vi}.

@@

FIXME again observe that up-smoothing is more efficient

@@

>>>>

During one application of \pr{nmcd-vcycle}, storage must be allocated for $\chi^j$, $\phi^j$, $g^j$, $y^j$, and $\ell^j$ on each level.  However, line 3 is essentially notational in that the finest-level iterate $g^J$ is the same as $w^J$; no copy is needed.  Similarly, in algorithm lines 13 and 15 no allocation is required because the $y^j$ space can be reused for $z^j$ as the coarser corrections are prolonged and accumulated.  Let $m^j$ be the number of interior nodes in the $j$th-level mesh.  The total storage of \pr{nmcd-vcycle}, excluding any in the smoothers/solver, but including for the original obstacle $\gamma^J$, is about $8 m^J$ real numbers\footnote{That is: $8\approx 5(4/3) + 1$.} under the standard estimate that a 2D, uniform-mesh V-cycle uses $4/3$ the memory of the finest level alone \cite[Section 2.4]{Trottenbergetal2001}.  For comparison, a single-level method needs at least $3 m^J$ storage for the vectors $\ell^J,\gamma^J,w^J$.

\pr{nmcd-vcycle} assumes certain semantics for the smoothers and the coarsest-level solver; the latter might also be a smoother algorithm.  They do in-place modifications of their final arguments, and they do not modify their first three arguments.  The smoothers are in-exact solvers of VIs \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi}, respectively, and they are iterated \id{down} and \id{up} times.  Further smoother implementation details and choices are addressed in the next section.

Repeated application of \pr{nmcd-vcycle} to solve VI \eqref{eq:fe:vi} will not reduce the finest-level residual $f^J(u^J) - \ell^J$ to zero everywhere.  However, recalling that a finite-dimensional VI is equivalent to a \emph{nonlinear complementarity problem} (NCP) \cite{FacchineiPang2003}, \eqref{eq:fe:vi} corresponds to the NCP
\begin{equation}
u^J \ge \gamma^J, \qquad f^J(u^J) \ge \ell^J, \qquad (u^J - \gamma^J)\left(f^J(u^J) - \ell^J\right) = 0.  \label{eq:fe:ncp}
\end{equation}
Thus, at convergence we only expect $f^J(u^J) = \ell^J$ at those locations where the constraint is inactive, namely where $u^J > \gamma^J$.  Therefore, recalling that the hat function at node $x_q^J$ is denoted $\psi_q^J$, for $w^J \in \mathcal{K}^J$ we define the following vector in $\RR^{m^J}$ as the finest-level \emph{nodal residual}:
\begin{equation}
\hat r(w^J)_q = \begin{cases} \ip{f^J(w^J)-\ell^J}{\psi_q^J}, & w^J(x_q^J) > \gamma^J(x_q^J), \\
                                  \min\{\ip{f^J(w^J)-\ell^J}{\psi_q^J},0\}, & w^J(x_q^J) = \gamma^J(x_q^J). \end{cases} \label{eq:cpresidual}
\end{equation}
(Formally, $\hat r:\mathcal{V}^J \to \RR^{m^J}$ is a nonlinear map.)  Observe that $\hat r(w^J)$ is nonzero if there are nodes where the NCP \eqref{eq:fe:ncp} is violated, either because $f^J(w^J)-\ell^J$ remains nonzero on the inactive set or because the same quantity is negative on the active set, and that nodal values of $f^J(w^J)-\ell^J$ are in evaluated in a weak sense, by application to a hat function.

A solver for \eqref{eq:fe:vi} using repeated applications of \pr{nmcd-vcycle} would start with an initial iterate $w^{J,0}$, do V-cycles, and stop once the finest-level nodal residual $\hat r(w^J)$ was sufficiently small.  For example, using tolerances \id{atol}$>0$ and \id{rtol}$>0$ the stopping criterion might be
\begin{equation}
\|\hat r(w^J)\| < \id{atol} \qquad \text{or} \qquad \frac{\|\hat r(w^J)\|}{\|\hat r(w^{J,0})\|} < \id{rtol},
\end{equation}
where $\|\cdot\|$ denotes a norm on $\RR^{m^J}$.  A step tolerance could also be used, measuring the difference between successive iterates.

@@

\section{Smoother implementations} \label{sec:smoothers}

FIXME

Recall that a $j$th-level hat function based at node $x_p^j$ is denoted $\psi_p^j$, for an iterate $g^j$ and an admissible correction $y^j \in \mathcal{D}^j$ we define the \emph{downward nodal residual} as
\begin{equation}
(\hat r_{\text{down}}(y^j))_p = \begin{cases} \ip{f^j(g^j+y^j)-\ell^j}{\psi_p^j}, & y^j(x_p^j) > \phi^j(x_p^j), \\
                                  \min\{\ip{f^j(g^j+y^j)-\ell^j}{\psi_p^j},0\}, & y^j(x_p^j) = \phi^j(x_p^j). \end{cases} \label{eq:cpresidual}
\end{equation}
The \emph{upward nodal residual} $\hat r_{\text{up}}(z^j)$ for $z^j \in \mathcal{U}^j$ is similarly defined using $\chi^j$.

FIXME stopping criteria for smoother

@@

FIXME observe that up-smoothing is more efficient
